# Gene Functional Enrichment Analysis

This report covers functional enrichment analysis of differentially expressed genes from *Escherichia coli* K-12 MG1655 after saccharin treatment ([Dedios *et al.*, 2025](https://link.springer.com/article/10.1038/s44321-025-00219-1)).

We apply two complementary approaches:

- **ORA** (Over-Representation Analysis) using [`mulea`](https://github.com/ELTEbioinformatics/mulea) — tests whether significant DE genes are enriched in KEGG pathways more than expected by chance, with empirical FDR correction
- **GSEA** (Gene Set Enrichment Analysis) using [`fgsea`](https://bioconductor.org/packages/fgsea/) — uses the full ranked gene list to detect coordinated pathway-level shifts, including subtle changes below the significance threshold

Both analyses use KEGG pathway gene sets retrieved via `KEGGREST` for *E. coli* K-12 MG1655.

### **Why two packages?**

:::boxy
 `mulea` provides excellent empirical FDR correction for ORA but its built-in GSEA
 function does not compute NES (Normalised Enrichment Scores). `fgsea` is the gold
 standard for preranked GSEA with proper NES and permutation-based p-values.
 Using the best tool for each task gives the most reliable results.
:::

### **Why KEGG?**

:::boxy
The original Dedios 2025 paper used KEGG pathway analysis for *E. coli* MG1655, making
this the biologically appropriate and directly comparable gene set database.
:::

### Load Libraries

```{r libraries-5, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(mulea)
library(fgsea)
library(KEGGREST)
library(knitr)
library(kableExtra)
library(DT)
library(plotly)
```


## Load Results from Differential Expression Analysis

This script reads the DE results produced by `02_differential_expression_ecoli.Rmd`.
Make sure you have run that script first — it saves two files to `results/`:

:::boxy
- `results/DE_treatment_vs_control_all.tsv` — all tested genes with shrunken LFC
- `results/DE_treatment_vs_control_significant.tsv` — significant DE genes (padj < 0.05, |LFC| >= 1)
:::

```{r load-results, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
git_root <- system("git rev-parse --show-toplevel", intern = TRUE)

res_df <- read.table(
  file.path(git_root, "results", "DE_treatment_vs_control_all.tsv"),
  header      = TRUE,
  sep         = "\t",
  check.names = TRUE
)

res_sig <- read.table(
  file.path(git_root, "results", "DE_treatment_vs_control_significant.tsv"),
  header      = TRUE,
  sep         = "\t",
  check.names = TRUE
)

cat("All tested genes     :", nrow(res_df), "\n")
cat("Significant DE genes :", nrow(res_sig), "\n")
cat("  Upregulated        :", sum(res_sig$log2FoldChange >= 1), "\n")
cat("  Downregulated      :", sum(res_sig$log2FoldChange <= -1), "\n")
```


## Build and Cache KEGG GMT

We retrieve KEGG pathway gene sets for *E. coli* K-12 MG1655 from the KEGG API using
`KEGGREST`. Both the GMT and the locus tag to gene symbol mapping table are cached locally
as RDS files so this step only runs once.

:::boxy
 **Why do we need a mapping step?**
 The nf-core/rnaseq pipeline uses locus tags (`b0001`, `b0002`...) as gene identifiers
 when no `gene_name` field is present in the reference GTF, which is common for
 prokaryotic genomes. The KEGG pathway GMT uses gene symbols (`aceE`, `recA`...).
 We build a locus tag to gene symbol map directly from KEGG and convert before analysis.
:::

<div style="background:#fff3cd;border-left:4px solid #ffc107;padding:10px;margin:10px 0;">
  <strong>Tip:</strong> The first time you run this chunk it queries the KEGG API (~2-3 min).
  After that both RDS files load instantly. Do not delete the files in
  <code>data/databases/</code>.
</div>

```{r build-kegg-gmt, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
gmt_path <- file.path(git_root, "data", "databases",
                      "KEGG_Pathways_Ecoli_MG1655_GeneSymbol.rds")
map_path <- file.path(git_root, "data", "databases",
                      "KEGG_locus_to_symbol_Ecoli_MG1655.rds")

if (file.exists(gmt_path) && file.exists(map_path)) {

  kegg_gmt <- readRDS(gmt_path)
  id_map   <- readRDS(map_path)
  cat("Loaded cached GMT    :", nrow(kegg_gmt), "pathways\n")
  cat("Loaded cached ID map :", nrow(id_map), "entries\n")

} else {

  cat("Fetching KEGG data for E. coli MG1655 — ~2-3 min...\n")

  ecoli_pathways <- keggList("pathway", "eco")
  pathway_ids    <- names(ecoli_pathways)
  pathway_names  <- gsub(" - Escherichia coli K-12 MG1655$", "",
                         as.character(ecoli_pathways))

  cat("Total pathways found:", length(pathway_ids), "\n")

  batches         <- split(pathway_ids, ceiling(seq_along(pathway_ids) / 10))
  pathway_genes   <- list()
  locus_to_symbol <- list()

  for (i in seq_along(batches)) {
    cat("  Batch", i, "of", length(batches), "\n")
    for (pid in batches[[i]]) {
      tryCatch({
        result <- keggGet(pid)[[1]]
        if (!is.null(result$GENE) && length(result$GENE) >= 2) {
          locus  <- result$GENE[seq(1, length(result$GENE), by = 2)]
          symbl  <- trimws(sub(";.*", "", result$GENE[seq(2, length(result$GENE), by = 2)]))
          pathway_genes[[pid]] <- symbl
          for (j in seq_along(locus)) locus_to_symbol[[locus[j]]] <- symbl[j]
        }
      }, error = function(e) cat("  Skipped:", pid, "\n"))
    }
    Sys.sleep(0.3)
  }

  kegg_gmt <- data.frame(
    ontology_id    = names(pathway_genes),
    ontology_name  = pathway_names[match(names(pathway_genes), pathway_ids)],
    list_of_values = I(pathway_genes),
    stringsAsFactors = FALSE
  )

  id_map <- data.frame(
    locus_tag   = names(locus_to_symbol),
    gene_symbol = unlist(locus_to_symbol),
    stringsAsFactors = FALSE
  )

  dir.create(dirname(gmt_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(kegg_gmt, gmt_path)
  saveRDS(id_map,   map_path)
  cat("GMT saved    :", nrow(kegg_gmt), "pathways\n")
  cat("ID map saved :", nrow(id_map), "entries\n")
}

cat("\nExample pathways:\n")
head(kegg_gmt$ontology_name, 6)
```


### Convert Locus Tags to Gene Symbols

```{r id-conversion, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
res_df_sym <- res_df %>%
  left_join(id_map, by = c("gene" = "locus_tag")) %>%
  mutate(gene = ifelse(!is.na(gene_symbol), gene_symbol, gene)) %>%
  select(-gene_symbol)

res_sig_sym <- res_sig %>%
  left_join(id_map, by = c("gene" = "locus_tag")) %>%
  mutate(gene = ifelse(!is.na(gene_symbol), gene_symbol, gene)) %>%
  select(-gene_symbol)

cat("Genes mapped to symbols  :", sum(res_df$gene %in% id_map$locus_tag),
    "of", nrow(res_df), "\n")
cat("Remaining as locus tags  :", sum(!res_df_sym$gene %in% id_map$gene_symbol),
    "(unannotated in KEGG - expected)\n")
cat("\nExample converted gene names:\n")
head(res_sig_sym$gene, 10)
```

<div style="background:#d1ecf1;border-left:4px solid #0c5460;padding:10px;margin:10px 0;">
  <strong>Note:</strong> Genes remaining as locus tags are not annotated in any KEGG
  pathway. They are included in the ORA background but will not match any gene set.
  This is expected — not all <em>E. coli</em> genes are assigned to KEGG pathways.
</div>

## Part 1 — Over-Representation Analysis (ORA)

**What is ORA?**

ORA asks: *"Are my significant DE genes enriched in any pathway more than expected by chance?"*

It uses a **hypergeometric test** (equivalent to a one-sided Fisher's exact test):

- **Query set** — significant DE genes (padj < 0.05, |LFC| >= 1), converted to gene symbols
- **Background** — all genes tested by DESeq2, converted to gene symbols
- **Gene sets** — KEGG pathways
- **FDR correction** — empirical FDR (eFDR) via resampling, which accounts for interdependence between gene sets

:::boxy
 **Key assumption:** ORA treats all significant genes equally — it ignores magnitude or
 direction of fold change. This is why we also run GSEA (Part 2), which uses the full
 ranked list.
:::


### Prepare Input

```{r ora-input, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
sig_genes <- res_sig_sym$gene
all_genes  <- res_df_sym$gene

cat("Query set (significant genes) :", length(sig_genes), "\n")
cat("Pool (all tested genes)        :", length(all_genes), "\n")
cat("Overlap with first pathway     :",
    length(intersect(sig_genes, kegg_gmt$list_of_values[[1]])), "genes\n")
```

<div style="background:#fff3cd;border-left:4px solid #ffc107;padding:10px;margin:10px 0;">
  <strong>Tip:</strong> Always use all <em>tested</em> genes as the background, not the full
  genome. DESeq2 already filtered to expressed genes, so <code>res_df_sym</code> is the
  correct pool. Using the full genome inflates the background and produces over-optimistic
  p-values.
</div>

### Run ORA

```{r ora-run, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
set.seed(42)

ora_model <- ora(
  gmt                       = kegg_gmt,
  element_names             = sig_genes,
  background_element_names  = all_genes,
  p_value_adjustment_method = "eFDR",
  number_of_permutations    = 1000
)

ora_results <- mulea::run_test(ora_model)

cat("Pathways tested          :", nrow(ora_results), "\n")
cat("Significant (eFDR < 0.05):", sum(ora_results$eFDR < 0.05, na.rm = TRUE), "\n")
```

### Filter and Inspect Significant Pathways

```{r ora-filter, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ora_sig <- ora_results %>%
  filter(eFDR < 0.05) %>%
  arrange(eFDR)

cat("Significant ORA pathways:", nrow(ora_sig), "\n")
```

```{r ora-table, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
DT::datatable(
  ora_sig %>%
    select(ontology_id, ontology_name, nr_common_with_tested_elements,
           nr_common_with_background_elements, p_value, eFDR) %>%
    mutate(across(where(is.numeric), \(x) round(x, 4))),
  rownames   = FALSE,
  colnames   = c("KEGG ID", "Pathway", "Hits in query",
                 "Hits in background", "p-value", "eFDR"),
  extensions = c("Buttons", "Scroller"),
  options    = list(
    dom      = "Bfrtip",
    buttons  = c("copy", "csv"),
    scrollX  = TRUE,
    scrollY  = 300,
    scroller = TRUE
  ),
  caption = "Significant ORA pathways — treatment vs control (eFDR < 0.05)"
)
```

**Visualise ORA Results**

### Bar Plot — Enriched Pathways

```{r ora-barplot, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ora_bar <- ora_sig %>%
  mutate(
    eFDR          = ifelse(eFDR == 0, 1e-10, eFDR),
    ontology_name = fct_reorder(ontology_name, -log10(eFDR))
  )

ora_barplot <- ggplot(ora_bar,
                      aes(x    = -log10(eFDR),
                          y    = ontology_name,
                          fill = -log10(eFDR))) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = -log10(0.05), linetype = "dashed",
             color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "#92C5DE", high = "#1A5276") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(
    title    = "ORA — Enriched KEGG Pathways (eFDR < 0.05)",
    subtitle = "Treatment vs Control | E. coli MG1655",
    x        = "-log10(eFDR)",
    y        = NULL
  )

ora_barplot
suppressMessages(ggplotly(ora_barplot))
```

### Lollipop Plot — Gene Hits per Pathway

```{r ora-lollipop, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ora_lollipop <- ggplot(ora_bar,
                       aes(x     = nr_common_with_tested_elements,
                           y     = ontology_name,
                           color = eFDR)) +
  geom_segment(aes(x    = 0,
                   xend = nr_common_with_tested_elements,
                   yend = ontology_name),
               linewidth = 0.8, color = "grey70") +
  geom_point(size = 4) +
  scale_color_gradient(low = "#CA0020", high = "#92C5DE", name = "eFDR") +
  theme_bw() +
  labs(
    title    = "ORA — Gene Hits per Pathway",
    subtitle = "Point colour = eFDR",
    x        = "Number of DE genes in pathway",
    y        = NULL
  )

ora_lollipop
suppressMessages(ggplotly(ora_lollipop))
```

### Critical Interpretation — Spurious Hits

:::boxy
**Always ask: does this pathway make biological sense for your organism?**

Two pathways in our results deserve scrutiny:

- **Pinene, camphor and geraniol degradation** (`eco00907`) — a plant terpenoid pathway.
  It appears here because KEGG annotates a handful of *E. coli* enzymes with broad
  substrate specificity to this pathway. The perfect overlap (6 hits in query, 6 in
  background, eFDR = 0) is a red flag: a tiny gene set where every background gene
  is DE will always score as maximally significant, regardless of biological relevance.

- **Caprolactam degradation** (`eco00930`) — same issue: only 3 genes in the background,
  all 3 in the query. Perfect overlap in a tiny set is statistically inevitable, not
  biologically meaningful.

**Thinking strategy:**
1. Check the gene set size — very small sets (< 5 genes) are prone to spurious enrichment
2. Check the overlap — 100% overlap is suspicious, not impressive
3. Ask whether the pathway is plausible for your organism and condition
4. Cross-reference with GSEA — if it does not appear there, treat it with scepticism
:::

```{r ora-filter-clean, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
# Remove biologically spurious pathways before downstream use
ora_sig_clean <- ora_sig %>%
  filter(!grepl("Pine|camphor|Caprolactam", ontology_name, ignore.case = TRUE))

cat("Pathways after removing spurious hits:", nrow(ora_sig_clean), "\n")
```

```{r ora-barplot-clean, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ora_bar_clean <- ora_sig_clean %>%
  mutate(
    eFDR          = ifelse(eFDR == 0, 1e-10, eFDR),
    ontology_name = fct_reorder(ontology_name, -log10(eFDR))
  )

ora_barplot_clean <- ggplot(ora_bar_clean,
                            aes(x    = -log10(eFDR),
                                y    = ontology_name,
                                fill = -log10(eFDR))) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = -log10(0.05), linetype = "dashed",
             color = "black", linewidth = 0.5) +
  scale_fill_gradient(low = "#92C5DE", high = "#1A5276") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(
    title    = "ORA — Enriched KEGG Pathways (eFDR < 0.05)",
    subtitle = "After removing spurious hits | E. coli MG1655",
    x        = "-log10(eFDR)",
    y        = NULL
  )

ora_barplot_clean
suppressMessages(ggplotly(ora_barplot_clean))
```

## Part 2 — Gene Set Enrichment Analysis (GSEA)

**What is GSEA?**

GSEA asks: *"Are genes in a pathway coordinately shifting up or down across my full ranked list?"*

Unlike ORA, GSEA:

- Uses **all tested genes** — capturing coordinated but subtle pathway-level changes
- **Ranks genes** by a metric encoding both significance and direction
- Returns a **Normalised Enrichment Score (NES)**: positive NES = pathway upregulated in treatment; negative NES = pathway downregulated

:::boxy
 **Why signed log p-value for ranking?**

 `rank = sign(LFC) x -log10(pvalue)`

 Genes at the top are significantly upregulated; genes at the bottom are significantly
 downregulated. This is more robust than ranking on LFC alone, which can place noisy
 high-fold-change genes from lowly expressed genes at the extremes.

<div class="note">
<strong>📘 Note:</strong> Genes with `pvalue = 0` (below machine precision in DESeq2) are excluded
 because `-log10(0) = Inf` is not a valid ranking score. These are the most extremely
 significant genes and their exclusion has minimal impact on pathway-level results.
</div>
:::

### Prepare the Ranked Gene List

```{r gsea-rank, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ranked_df <- res_df_sym %>%
  filter(!is.na(pvalue), !is.na(log2FoldChange)) %>%
  filter(pvalue > 0) %>%                    # exclude p=0 -> log10(0) = Inf
  mutate(rank_metric = sign(log2FoldChange) * -log10(pvalue)) %>%
  arrange(desc(rank_metric)) %>%
  distinct(gene, .keep_all = TRUE)

rank_vector        <- ranked_df$rank_metric
names(rank_vector) <- ranked_df$gene

cat("Genes in ranked list :", length(rank_vector), "\n")
cat("Infinite values      :", sum(!is.finite(rank_vector)), "\n")
cat("Top 5 (upregulated)  :", paste(head(names(rank_vector), 5), collapse = ", "), "\n")
cat("Bottom 5 (downreg.)  :", paste(tail(names(rank_vector), 5), collapse = ", "), "\n")
```

### Convert GMT to fgsea Format

`fgsea` requires a named list of gene vectors — one entry per pathway.

```{r gsea-pathways, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
pathways_list <- setNames(
  kegg_gmt$list_of_values,
  kegg_gmt$ontology_name
)

cat("Pathways in list:", length(pathways_list), "\n")
cat("Example entry   :", pathways_list[[1]][1:5], "\n")
```

### Run GSEA

```{r gsea-run, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
set.seed(42)

gsea_results <- fgsea(
  pathways    = pathways_list,
  stats       = rank_vector,
  minSize     = 10,
  maxSize     = 500,
  nPermSimple = 10000
)

cat("Pathways tested          :", nrow(gsea_results), "\n")
cat("Significant (padj < 0.05):", sum(gsea_results$padj < 0.05, na.rm = TRUE), "\n")
cat("Nominal (pval < 0.05)    :", sum(gsea_results$pval < 0.05, na.rm = TRUE), "\n")
```

### Filter and Inspect Significant Pathways

```{r gsea-filter, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
gsea_sig <- gsea_results %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>%
  arrange(padj)

cat("Significant GSEA pathways:", nrow(gsea_sig), "\n")
cat("  Activated (NES > 0)    :", sum(gsea_sig$NES > 0), "\n")
cat("  Suppressed (NES < 0)   :", sum(gsea_sig$NES < 0), "\n")
```

```{r gsea-table, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
DT::datatable(
  gsea_sig %>%
    select(pathway, NES, pval, padj, size) %>%
    mutate(
      direction = ifelse(NES > 0, "Activated", "Suppressed"),
      across(where(is.numeric), \(x) round(x, 4))
    ) %>%
    arrange(desc(abs(NES))),
  rownames   = FALSE,
  colnames   = c("Pathway", "NES", "p-value", "padj", "Size", "Direction"),
  extensions = c("Buttons", "Scroller"),
  options    = list(
    dom      = "Bfrtip",
    buttons  = c("copy", "csv"),
    scrollX  = TRUE,
    scrollY  = 300,
    scroller = TRUE
  ),
  caption = "Significant GSEA pathways — treatment vs control (padj < 0.05)"
)
```

**Visualise GSEA Results**

### NES Bar Plot — Activated and Suppressed Pathways

```{r gsea-nes-barplot, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
cols_direction <- c("Activated" = "#d7191c", "Suppressed" = "#2c7bb6")

# Use top 20 by |NES| if more than 20 significant, otherwise show all
gsea_bar <- gsea_results %>%
  as.data.frame() %>%
  filter(pval < 0.05) %>%
  slice_max(abs(NES), n = 20) %>%
  mutate(
    direction = ifelse(NES > 0, "Activated", "Suppressed"),
    pathway   = fct_reorder(pathway, NES)
  )

gsea_nesplot <- ggplot(gsea_bar,
                       aes(x    = NES,
                           y    = pathway,
                           fill = direction)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed",
             color = "grey50", linewidth = 0.4) +
  scale_fill_manual(values = cols_direction) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(
    title    = "GSEA — Normalised Enrichment Scores (pval < 0.05)",
    subtitle = "Treatment vs Control | E. coli MG1655",
    x        = "Normalised Enrichment Score (NES)",
    y        = NULL,
    fill     = NULL
  )

gsea_nesplot
ggplotly(gsea_nesplot)
```

### Dot Plot — NES, padj and Gene Set Size

```{r gsea-dotplot, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
gsea_dotplot <- ggplot(gsea_bar,
                       aes(x     = NES,
                           y     = pathway,
                           size  = size,
                           color = padj)) +
  geom_point(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_gradient(low = "#CA0020", high = "#92C5DE", name = "padj") +
  scale_size_continuous(name = "Genes in set", range = c(3, 10)) +
  theme_bw() +
  theme(legend.position = "right") +
  labs(
    title    = "GSEA — Pathway Dot Plot",
    subtitle = "Point size = genes in set | Colour = padj",
    x        = "NES",
    y        = NULL
  )

gsea_dotplot
ggplotly(gsea_dotplot)
```

### Enrichment Plot — Top Pathways

The enrichment plot shows the running sum of the enrichment score for a single pathway
across the ranked gene list. The peak of the curve is the enrichment score. Genes from
the pathway are shown as vertical bars (the "rug") at the bottom.

```{r gsea-enrichment-plots, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
# Top activated pathway
top_up <- gsea_results %>%
  as.data.frame() %>%
  filter(NES > 0) %>%
  slice_min(pval, n = 1) %>%
  pull(pathway)

# Top suppressed pathway
top_down <- gsea_results %>%
  as.data.frame() %>%
  filter(NES < 0) %>%
  slice_min(pval, n = 1) %>%
  pull(pathway)

if (length(top_up) > 0) {
  plotEnrichment(pathways_list[[top_up]], rank_vector) +
    labs(title = paste("Top activated:", top_up))
}

if (length(top_down) > 0) {
  plotEnrichment(pathways_list[[top_down]], rank_vector) +
    labs(title = paste("Top suppressed:", top_down))
}
```


## Part 3 — Comparing ORA and GSEA

ORA and GSEA are complementary — they answer different questions and will not always agree.

```{r compare, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
ora_hits  <- ora_sig$ontology_name
gsea_hits <- gsea_sig$pathway

both      <- intersect(ora_hits, gsea_hits)
ora_only  <- setdiff(ora_hits, gsea_hits)
gsea_only <- setdiff(gsea_hits, ora_hits)

cat("Pathways significant in BOTH :", length(both), "\n")
cat("Pathways in ORA only         :", length(ora_only), "\n")
cat("Pathways in GSEA only        :", length(gsea_only), "\n")

if (length(both) > 0) {
  cat("\nPathways found by both methods:\n")
  cat(paste(" -", both), sep = "\n")
}
```

<div style="background:#d4edda;border-left:4px solid #28a745;padding:10px;margin:10px 0;">
  <strong>Interpretation guide:</strong>
  <ul>
    <li><strong>Both significant</strong> — high confidence; strong pathway-level change
    driven by many DE genes</li>
    <li><strong>ORA only</strong> — genes crossing the significance threshold cluster in
    this pathway, but the overall coordinated shift may be modest</li>
    <li><strong>GSEA only</strong> — coordinated but subtle change below the LFC/padj
    threshold; genes shift together without individual genes being significant</li>
  </ul>
</div>



## Saving Results

```{r save-results-5, echo=TRUE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE}
results_dir <- file.path(git_root, "results")
dir.create(results_dir, showWarnings = FALSE)

write.table(
  ora_results,
  file      = file.path(results_dir, "ORA_KEGG_treatment_vs_control.tsv"),
  sep       = "\t",
  quote     = FALSE,
  row.names = FALSE
)

write.table(
  as.data.frame(gsea_results) %>% select(-leadingEdge),
  file      = file.path(results_dir, "GSEA_KEGG_treatment_vs_control.tsv"),
  sep       = "\t",
  quote     = FALSE,
  row.names = FALSE
)

cat("Results saved to:", results_dir, "\n")
cat("  ORA  -> ORA_KEGG_treatment_vs_control.tsv\n")
cat("  GSEA -> GSEA_KEGG_treatment_vs_control.tsv\n")
```

## Summary

| Step | Choice made | Rationale |
|---|---|---|
| ORA package | `mulea` | Empirical FDR correction; accounts for gene set interdependence |
| GSEA package | `fgsea` | Gold standard for preranked GSEA; proper NES and permutation p-values |
| Gene sets | KEGG pathways | Matches Dedios 2025; biologically interpretable for bacteria |
| GMT source | `KEGGREST` API + local RDS cache | No pre-built KEGG GMT for *E. coli*; built from API and cached |
| ID conversion | Locus tag to gene symbol via KEGGREST | nf-core uses locus tags when GTF lacks gene_name; KEGG GMT uses symbols |
| ORA background | All DESeq2-tested genes | Correct statistical background — not the full genome |
| GSEA rank metric | sign(LFC) x -log10(pvalue) | Encodes direction and significance; robust to lowly expressed gene noise |
| ORA FDR | Empirical FDR (eFDR) | Accounts for gene set interdependence |
| GSEA FDR | Benjamini-Hochberg (fgsea default) | Standard for preranked GSEA |
| Significance threshold | 0.05 | Standard; consistent with DE analysis |

Enrichment results are saved and ready for biological interpretation and reporting.


<!-- --------------------- Do not edit this and below ---------------------- -->
</br>

```{r session-info6, echo=TRUE, eval=TRUE, include=TRUE}
sessionInfo()
```

```{r , eval=FALSE, echo=FALSE}
rmarkdown::render("03_gene_functional_ecoli.Rmd")
```
