---
title: "Introduction to Transcriptomics"
subtitle: "Workshop material"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook
---
# Welcome
 
```{r,echo=FALSE,include=FALSE}
#  variables
rep_org <- "Data Science Platform" 
rep_core  <- "Juliana Assis"
rep_core_email  <- "jasge@dtu.dk"
rep_core2 <- "Sebastian Schulz"
rep_core_email2 <- "sebschu@dtu.dk"
rep_core3 <- "Alberto Pallejà Caro"
rep_core_email3 <- "apca@dtu.dk"
rep_ta    <- "Edir Sebastian Vidal Castro"
rep_ta_email <- "s243564@student.dtu.dk"

# Combine all responsibles into a single string
rep_all <- paste0(
  rep_core, " (", rep_core_email, ")<br>",
  rep_core2, " (", rep_core_email2, ")<br>",
  rep_core3, " (", rep_core_email3, ")"
)

rep_ta_full <- paste0(rep_ta, " (", rep_ta_email, ")")
```

<br>

::: boxy
__Organisation:__ `r rep_org`  
__Responsibles:__ `r paste0(rep_all," ")`  
__Teaching Assistant:__ `r rep_ta_full`  
:::

<br>


**Welcome to the Introduction to Transcriptomics Workshop!**

This workshop introduces **RNA-seq data analysis**, covering both the theory behind the technology and the full practical workflow — from raw reads to biological interpretation — using real *E. coli* data.

The course is designed for people with **no or little knowledge and practical experience** in transcriptomic analysis using RNA sequencing (RNAseq).

**Objectives**

* Theory on Illumina short-read sequencing technology for transcriptome analysis
* Theory and practical sessions for
   * RNAseq data processing using the nf-core/rnaseq pipeline
   * Statistical data analysis and visualization of RNAseq results
   * Enrichment analysis — gaining biological insights from transcriptomic data
   * Effective results reporting using the in-house developed software **[vuegen](https://github.com/Multiomics-Analytics-Group/vuegen)**

:::boxy
You can also explore the [DSP Transcriptomics Training repository](https://github.com/biosustain/dsp_transcriptomics_training)
:::

| Time | Session |
|---|---|
| ☕ **9:00–9:15** | Welcome, coffee, and setup |
| 🔬 **9:15–10:00** | Theory: Illumina sequencing & RNA-seq experimental design |
| 🔄 **10:00–10:45** | Theory: nf-core/rnaseq pipeline overview |
| ⏸️ **10:45–11:00** | Break |
| 💻 **11:00–12:00** | Script 01: *nf-core/rnaseq — running the pipeline* |
| 🍽️ **12:00–13:00** | Lunch |
| 🧬 **13:00–14:15** | Script 02: *Quality Control & Exploratory Data Analysis* |
| ⏸️ **14:15–14:30** | Break |
| 📊 **14:30–15:30** | Script 03: *Differential Expression Analysis with DESeq2* |
| ⏸️ **15:30–15:45** | Break |
| 🔍 **15:45–16:15** | Script 04: *Functional Enrichment Analysis (ORA & GSEA with mulea)* |
| 🏁 **16:15–16:30** | Wrap-up, Q&A, and closing |


:::boxy
Material for the workshop is located at:
[dsp_transcriptomics_training](https://github.com/biosustain/dsp_transcriptomics_training)
:::

Below are two setup options for the practical activities:

>Run the workshop in the cloud (no need to install anything). 

**Launch the app:**

[![Open in GitHub Codespaces](https://github.com/codespaces/badge.svg)](https://codespaces.new/biosustain/dsp_transcriptomics_training)


>Run the workshop locally on your machine. 

To run it on your own machine, install the following packages:

```{r, eval=FALSE}
# Install BiocManager if needed
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Bioconductor packages
BiocManager::install(c(
  "DESeq2",
  "SummarizedExperiment",
  "apeglm",
  "EnhancedVolcano"
))

# CRAN packages
packages <- c(
  "dplyr",
  "ggplot2",
  "ggrepel",
  "RColorBrewer",
  "rmarkdown",
  "tidyr",
  "plotly",
  "pheatmap",
  "ggpubr",
  "here",
  "mulea"
)

install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org/")
  } else {
    message(pkg, " already installed.")
  }
}

invisible(lapply(packages, install_if_missing))

message("All requested R packages are installed and ready to use!")
```
